Crunching Large Graphs with Commodity Processors

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
fix spacings and cache/net figs move to top
some modifications

%\paragraph{Multi-node ideas.}
As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
interact, so that threads waiting on long-latency memory operations or
synchronization operations will be scheduled soon after their
operations have completed.

@@

%\paragraph{Single-node implementation.}
In our single-node implementation, we use prefetch instructions for the issue operation, and regular
blocking reads and writes for the complete operation.  Since we expect
these global references to have low locality, we use the
``non-temporal'' form of prefetch to avoid disturbing the state of the
cache hierarchy \cite{intel:swdev}.

@@

With either approach, the question of request aggregation will be
important. Each memory request we make is small: 8 bytes is a likely
size. But most networks are optimized for bulk transfers of a few
thousand bytes. To improve network utilization, we will explore the
aggregation of multiple memory operations to different addresses on
the same remote node.

@@

\begin{figure*}[ht]
\vspace{-0.2in}
%list walk ILP
\begin{minipage}[b]{0.3\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
    list1 = list1->next;
    list2 = list2->next;
    \ldots
    list\(n\) = list\(n\)->next;
  \}
  }\centering{\bf (a)}
\end{alltt}
%\caption{Pseudocode for pointer chasing without coroutines.}
\label{fig:pointernocoro}
}\end{minipage}
%listwalk green threads
\begin{minipage}[b]{0.35\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
     readIssue(&(list->next));
     yield();
     list = readCompete(&(list->next));
 \}
 }\centering{\bf (b)}
\end{alltt}
  %%n lists
  % while (count-- > 0) \{
  % prefetch(&(list1->next));
  %   prefetch(&(list2->next));
   %  \ldots
   %  prefetch(&(list\(n\)->next));
   %  switch();
   %  list1 = read(&(list1->next));
   %  list2 = read(&(list2->next));
   %  \ldots
   %  list\(n\) = read(&(list\(n\)->next));
%\caption{Pseudocode for pointer chasing using coroutines.}
\label{fig:pointercoro}
}\end{minipage}
%cache pressure
\begin{minipage}[b]{0.32\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
    readIssue(&(remoteList->next));
    yield();
    remoteList = readComplete(&(remoteList->next));
    for( i in 1 to num_local_updates ) \{
      localList->data++;
      localList = localList->next;
   \} \}
   }\centering{\bf (c)}
\end{alltt}
%\caption{Pseudocode for pointer chasing with local updates.}
\label{fig:pointerupdate}
}\end{minipage}
\vspace{5pt}
\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates. Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
\label{fig:code}
\end{figure*}

@@

Figure~\ref{fig:multi-green} shows the result. For this
experiment, we limited memory concurrency to one concurrent miss per
coroutine, so the only source of memory concurrency is the use of
coroutines. We are able to obtain a rate of \mrps{275} with 48 concurrent misses, or 8 coroutines per core. These
results are very close to the ILP-only experiment, but more concurrent requests are required--about 48 versus 36. When the number of references per coroutine is larger than 1, fewer coroutines are required to reach full performance.

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In the multi-node runtime, \texttt{readIssue} will not be implemented with a prefetch instruction. The single-node coroutine library does depend on prefetches: other experiments show that without them, the overhead of context switches limits concurrency when there are more than two coroutines per core. Lastly, we observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per core. 
With 1 to 4 updates per remote reference for smaller working sets, the runtime achieves near maximum throughput. As working set size increases, cache misses in the local space become a bottleneck. When there are several updates and large working sets, the data of other coroutines is likely to be evicted from cache and performance suffers.

@@

To evaluate the potential for the lightweight contexts to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.

Since the network interface is accessed over QPI, we estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine. To do this, we allocated the lists in the second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to \mrps{175}.

@@

There has also been much work on user level threading, including the First-Class
User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
different goal \todo{we didnt say what their goal is}; we want many lightweight contexts to tolerate memory
latency.

>>>>

%\paragraph{Multi-node ideas.}
%As we develop our multi-node implementation, we will explore how the memory manager and coroutine scheduler may
%interact, so that threads waiting on long-latency memory operations or
%synchronization operations will be scheduled soon after their
%operations have completed.
As we develop our multi-node implementation, we will improve our
scheduler to activate threads after their long-latency operations
have completed.

% will explore how the memory manager and coroutine scheduler may
%interact, so that threads waiting on long-latency memory operations or
%synchronization operations will be scheduled soon after their
%operations have completed.

@@

%\paragraph{Single-node implementation.}
In our single-node implementation, we use prefetch instructions for the issue operation, and regular
blocking reads and writes for the complete operation.
%  Since we expect
%these global references to have low locality, we use the
%``non-temporal'' form of prefetch to avoid disturbing the state of the
%cache hierarchy \cite{intel:swdev}.

@@

With either approach, the question of request aggregation will be
important. Each memory request we make is small---perhaps 8
bytes---but most networks are optimized for bulk transfers of a few
thousand bytes. To improve network utilization, we will explore the
aggregation of multiple memory operations to different addresses on
the same remote node.

@@

\begin{figure*}[ht]
\vspace{-0.2in}
%list walk ILP
\begin{minipage}[b]{0.3\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
    list1 = list1->next;
    list2 = list2->next;
    \ldots
    list\(n\) = list\(n\)->next;
  \}
  }\centering{\bf (a)}
\end{alltt}
%\caption{Pseudocode for pointer chasing without coroutines.}
\label{fig:pointernocoro}
}\end{minipage}
%listwalk green threads
\begin{minipage}[b]{0.35\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
     readIssue(&(list->next));
     yield();
     list = readComplete(&(list->next));
 \}
 }\centering{\bf (b)}
\end{alltt}
  %%n lists
  % while (count-- > 0) \{
  % prefetch(&(list1->next));
  %   prefetch(&(list2->next));
   %  \ldots
   %  prefetch(&(list\(n\)->next));
   %  switch();
   %  list1 = read(&(list1->next));
   %  list2 = read(&(list2->next));
   %  \ldots
   %  list\(n\) = read(&(list\(n\)->next));
%\caption{Pseudocode for pointer chasing using coroutines.}
\label{fig:pointercoro}
}\end{minipage}
%cache pressure
\begin{minipage}[b]{0.32\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
    readIssue(&(remoteList->next));
    yield();
    remoteList = readComplete(&(remoteList->next));
    for( i in 1 to num_local_updates ) \{
      localList->data++;
      localList = localList->next;
   \} \}
   }\centering{\bf (c)}
\end{alltt}
%\caption{Pseudocode for pointer chasing with local updates.}
\label{fig:pointerupdate}
}\end{minipage}
\vspace{5pt}
\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates. Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
\label{fig:code}
\end{figure*}

@@

Figure~\ref{fig:multi-green} shows the result. For this experiment,
the only source of memory concurrency is the use of coroutines. We are
able to obtain a rate of \mrps{275} with 48 concurrent misses, or 8
coroutines per core. More concurrent requests are required to saturate
memory bandwidth than in the ILP-only experiment.

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In
the multi-node runtime, \texttt{readIssue} will not be implemented
with a prefetch instruction. The single-node coroutine library does
depend on prefetches: further experiments show that without them, the overhead of context switches limits concurrency when there are more than two coroutines per core. Lastly, we observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per
core.  With 1 to 4 updates per remote reference for smaller working
sets, the runtime achieves near maximum throughput. As working set
size increases, cache misses in the local space become a
bottleneck. When there are several updates to large working sets, the
data of other coroutines is likely to be evicted from cache and
performance suffers.

@@

To evaluate the potential for the lightweight contexts to perform in
the multi-node case, we simulated a network delay to see whether the
runtime could tolerate the larger latency.

Any network communication must travel over the QPI link in our test
system. We estimated the bandwidth of this link by allocating lists in
the first chip's memory and traversing those lists on the second
chip's cores. We found that use of the QPI link limits us to
\mrps{175}.

@@

There has also been much work on user level threading, including the First-Class
User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
different goal; we want many lightweight contexts to tolerate memory
latency.
